On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-03 manuscripts/PNAS/review/point_by_point_response.txt

RESPONSE: We have included in our revised manuscript a more compelling exposition of the significance of our results and their relevance to the general scientific audience, and thank the editor for the opportunity to address this concern shared by both reviewers. [Summarize high level major points in our cover here.  Also, mention we try to flesh this out in the next comment below.]

-----------------	
Since the result is an extension of the noise-free result of [Hillar & Sommer (2015)] to a noisy case, the authors should provide better justification for the significance of contributions. The main result seems to be based on an approach similar to [Hillar & Sommer (2015)], and there is little to no discussion of the computability and tightness of the sufficient condition for stability in the main results. Thus, the submission does not seem to be greatly more valuable than [Hillar & Sommer (2015)] and also not conclusive regarding the central problems of sparse linear coding and dictionary learning. That being said, the paper (with the modifications below) would certainly merit publication in very solid journal like IEEE Transactions on Information Theory, where it would be a median paper, in my opinion.  
-----------------

RESPONSE:

We agree with the reviewer that a straight-forward extension of the approach and noise-free result of [Hillar & Sommer (2015)] to the noisy case — while still a significant advancement in applied mathematics with actual practical implications — would not necessarily merit publication in PNAS on its own. The reviewer, however, seems to have overlooked the fact that our main result (Thm. 1) generalizes the noise-free result of [Hillar & Sommer (2015)] far beyond a simple extension to the noisy case in ways which provide valuable insight into the dictionary learning model. No explicit comparison was made in the manuscript, but a significant deviation from the approach taken in [Hillar & Sommer (2015)] was necessary to prove these new facts. Moreover, our solution (Cor. 2) to the optimization formulation of the dictionary learning problem (Prob. 2) — that of most interest to optimization theorists and practitioners of their methods — is the first of its kind in both the noise-free and noisy domains. We discuss all of these points in more detail in our response to the reviewer’s later “detailed comments”.

We have included in our updated manuscript a discussion of the computability of the sufficient conditions (they are, indeed, computable), as per the reviewer’s request. It is true, however, that we have not proven the tightness of these conditions and the manuscript is therefore not *conclusive* regarding the central problems of sparse linear coding. We believe that these sufficient conditions yield strong enough conclusions to stand on their own. Nonetheless, we have included in our revised manuscript some discussion of where the results can be tightened (e.g. Lem. 4). As for tightness of the sample complexity N, we believe that our N is on the order of the smallest possible given a lack of any additional assumptions *at all* on the reproducing matrix B and sparse codes \tilde x_i. A sample complexity polynomial in (m,k) can be achieved, however, by limiting the size of the support set of the \tilde x_i, and we have revised the manuscript to discuss this fact (which, incidentally, does *not* follow from the results of [Hillar & Sommer (2015)] in the noiseless case).

@@

-- The paper does a good job of motivating the general problem of sparse coding to a general audience. In particular, there is a good review of the areas in which dictionary learning arises.

-- I think the biggest potential issue for publication in PNAS is that the significance of the result needs more discussion. Specifically, the authors could address why going from the noise-free case to the noisy case is not incremental. PNAS specifically targets submissions that go beyond what could be published in a more specialized journal. Could the authors, for example, contrast the proof technique between the noisy and noise-free cases? Or, perhaps, the practical implications? 
-----------------

RESPONSE:

With all due respect to the reviewer, we believe the most compelling practical implication of a noisy result over a noiseless result is plainly evident: never has there been an experiment without noise and an unstable model is therefore useless in practice. We assume the reviewer must instead be concerned with what additional practical “take-away” message our manuscript contains. In brief, it is this: there exists a procedure by which one can determine if one’s proposed solution to Prob. 1 or Prob. 2 is indeed unique. We have updated the manuscript with an explicit statement of this fact.

The reviewer also seems concerned that, regardless of practical implications, our proof may amount to too straight-forward of an extension of the approach taken in [Hillar & Sommer (2015)] to satisfy the criterion for publication in PNAS. We believe this concern stems from two misconceptions of the reviewer: that our main result (Thm. 1) is merely an extension of that in [Hillar & Sommer (2015)] to the noisy case, and that our proof technique is only an incremental tweak on the approach taken in [Hillar & Sommer (2015)].

It is understandable from the presentation of our proof of our main result (Thm. 1) that our approach appears on the surface to be a straight-forward extension of that taken in [Hillar & Sommer (2015)]. As observed by our second reviewer, however, the extension to the noisy case did indeed require an original combination of several results in the literature. Aside from these required modifications, our proof of (Thm. 1) differs from the approach taken by [Hillar & Sommer (2015)] in a key way that is tucked into the proof of our Main Lemma — more specifically, by the addition of Lemma 5. [More detail on what goes down in Lemma 5?] We have updated the manuscript to make the reader more keenly aware of this. To be clear, the perspective we take on the problem yields the following powerful new conclusions beyond those of a straight-forward extension of the noiseless result of [Hillar & Sommer (2015)] to the noisy case:

### 2017-08-03 manuscripts/PNAS/chazthm_pnas.tex
chaz

% TODO %
% *** Put ell1 norm into the denominator of L_k *** ?
% Give example of when we might expect k-uniformity, i.e. x_i in glp on k-dimensional subspaces. Orbit under group action of a k-dimensional subspace?
% Proof of cor 2 can do sum starting with p=1 since they are in general linear position, hence none can be 0. but whatever.

@@

\begin{definition}\label{dDef}
For $\mathbf{u} \in \mathbb R^m$ and vector spaces $U,V \subseteq \mathbb{R}^m$, let $\text{\rm dist}(\mathbf{u}, V) := \inf \{\| \mathbf{u}-\mathbf{v} \|_2: \mathbf{v} \in V\}$ and define:
\begin{align}\label{d}
d(U,V) := \sup_{\mathbf{u} \in U, \ \|\mathbf{u}\|_2 \leq 1} \text{\rm dist}(\mathbf{u},V).
\end{align}
\end{definition}

@@

\section{Discussion}\label{Discussion}
%[mention dictionary size independence]
In this note, we generalized the approach of \cite{Hillar15} to prove the stability of unique solutions to Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} while significantly reducing the known sample complexity. Our results justify the application of the sparse linear coding model to blind source separation problems, wherein the goal is to infer the generating dictionary and sparse codes from noisy measurements, \eqref{LinearModel}. We also collect a set of useful mathematical tools and basic facts for future research. The main motivation for this work, however, was to understand how seemingly universal representations emerge from sparse coding models fit to natural data by a variety of methodologies. We elaborate on these applications below after discussing theoretical aspects.

@@

One notable component of our contribution is a combinatorial criteria (regular hypergraphs satisfying the singleton intersection property, Def.~\ref{sip}) for the support sets of sparse codes key to the identification of the dictionary. Fully understanding those combinatorial designs allowing for stable sparse representations is an interesting research area for the future. For instance, whether there is a support set size $N$ that is polynomial in $m,k$ gauranteeing the conclusions of Thm.~\ref{DeterministicUniquenessTheorem} has implications for the computational complexity of Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} and related questions \cite{Tillmann15}.

A technical difficulty in proving Thm.~\ref{DeterministicUniquenessTheorem} was the absence of a spark condition assumption on solutions to Prob.~\ref{InverseProblem}. Although mathematically interesting that no such requirement is necessary, there are other reasons to seek out such a theoretical guarantee. For instance, it is difficult to ensure that an algorithm maintain a dictionary satisfying \eqref{SparkCondition} at each iteration; indeed, even certifying a dictionary has this property is likely intractable given its NP-hardness \cite{tillmann2014computational}.

@@

We begin by showing that:
\begin{align}\label{eq2}
\dim(\bm{\mathcal{B}}_{\pi(S)}) = \dim(\bm{\mathcal{A}}_S) \ \ \text{for all $S \in \mathcal{H}$}. 
\end{align}
Since the right-hand side of \eqref{GapUpperBound} is less than one (by \eqref{delrho} and $r \leq 1$), it follows from \eqref{dimLem} that $|\pi(S)| \geq \dim(\bm{\mathcal{B}}_{\pi(S)}) \geq \dim(\bm{\mathcal{A}}_S) = |S|$, with the equality by injectivity of $\mathbf{A}$. Since $|S| = |\pi(S)|$, the required fact follows. Note that the columns of $\mathbf{B}_{\pi(S)}$ are therefore linearly independent, for all $S \in \mathcal{H}$.

We next show that \eqref{cond} holds.  Fix $\mathcal{G} \subseteq \mathcal{H}$. Since $\bm{\mathcal{B}}_{\cap \pi(\mathcal{G})} \subseteq \cap_{S \in \mathcal{G}} \bm{\mathcal{B}}_{\pi(S)}$, if $\cap_{S \in \mathcal{G}} \bm{\mathcal{B}}_{\pi(S)} = \{\textbf{0}\}$, then we must have $|\cap_{S \in \mathcal{G}} \pi(S)| = 0$ (as the columns of $\mathbf{B}_{\pi(S)}$ are linearly independent) and \eqref{cond} is trivially true. Suppose then that the intersection is not the zero vector. Applying (in order) Lem.~\ref{eqdim}, Lem.~\ref{DistanceToIntersectionLemma}, \eqref{dDef}, \eqref{eq2} and finally \eqref{GapUpperBound}, we have:
\begin{align}\label{randoml}
d(\bm{\mathcal{B}}_{\cap \pi(\mathcal{G})}&, \bm{\mathcal{A}}_{\cap \mathcal{G}}  ) 
\leq d\left( \cap_{S \in \mathcal{G}} \bm{\mathcal{B}}_{\pi(S)}, \cap_{S \in \mathcal{G}} \bm{\mathcal{A}}_S \right) \nonumber \\
&\leq \sum_{T \in \mathcal{G}} \frac{ d\left( \cap_{S \in \mathcal{G}} \bm{\mathcal{B}}_{\pi(S)},\bm{\mathcal{A}_{T}} \right) }{ r(  \{\bm{\mathcal{A}}_S\}_{S \in \mathcal{G}}) }
\leq \sum_{T \in \mathcal{G}} \frac{ d\left( \bm{\mathcal{B}}_{\pi(T)},\bm{\mathbf{A}}_{T} \right) }{ r( \{\bm{\mathcal{A}}_S\}_{S \in \mathcal{G}}) }\nonumber \\
&\leq \frac{|\mathcal{G}| \varepsilon}{r( \{\bm{\mathcal{A}}_S\}_{S \in \mathcal{G}})} 
\leq \frac{C_2 \varepsilon}{\max_i\|\mathbf{A}_i\|_2}. 
\end{align}

>>>>

% TODO %
% *** Put ell1 norm into the denominator of L_k *** ?
% Give example of when we might expect k-uniformity, i.e. x_i in glp on k-dimensional subspaces. Orbit under group action of a k-dimensional subspace?

@@

\begin{definition}\label{dDef}
For $\mathbf{u} \in \mathbb R^m$ and vector spaces $U,V \subseteq \mathbb{R}^m$, let $\text{\rm dist}(\mathbf{u}, V) := \inf \{\| \mathbf{u}-\mathbf{v} \|_2: \mathbf{v} \in V\}$ and define:
\begin{align}
d(U,V) := \sup_{\mathbf{u} \in U, \ \|\mathbf{u}\|_2 \leq 1} \text{\rm dist}(\mathbf{u},V).
\end{align}
\end{definition}

@@

\section{Discussion}\label{Discussion}
%[mention dictionary size independence]

% A significant implication is that the constant $C_1$ of Thm.~\ref{DeterministicUniquenessTheorem} can be computed efficiently if we restrict the class of possible supports for the $\mathbf{\bar x}_i$.

In this note, we generalized the approach of \cite{Hillar15} to prove the stability of unique solutions to Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} while significantly reducing the known sample complexity. Our results justify the application of the sparse linear coding model to blind source separation problems, wherein the goal is to infer the generating dictionary and sparse codes from noisy measurements, \eqref{LinearModel}. We also collect a set of useful mathematical tools and basic facts for future research. The main motivation for this work, however, was to understand how seemingly universal representations emerge from sparse coding models fit to natural data by a variety of methodologies. We elaborate on these applications below after discussing theoretical aspects.

@@

One notable component of our contribution is a combinatorial criteria (regular hypergraphs satisfying the singleton intersection property, Def.~\ref{sip}) for the support sets of sparse codes key to the identification of the dictionary. Fully understanding those combinatorial designs allowing for stable sparse representations is an interesting research area for the future. For instance, whether there is a sample size $N$ that is polynomial in $m,k$ gauranteeing the conclusions of Thm.~\ref{DeterministicUniquenessTheorem} has implications for the computational complexity of Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} and related questions \cite{Tillmann15}. A re-examination of pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} demonstrates that this can in fact be achieved if one restricts the supports of the $\mathbf{\bar x}_i$ to a class that grows polynomially in $m, k$.

A technical difficulty in proving Thm.~\ref{DeterministicUniquenessTheorem} was the absence of any kind of assumption on solutions to Prob.~\ref{InverseProblem}. Although mathematically interesting that no such requirement is necessary, there are other reasons to seek out such a theoretical guarantee. For instance, it is difficult to ensure that an algorithm maintain a dictionary satisfying \eqref{SparkCondition} at each iteration; indeed, even certifying a dictionary has this property is likely intractable given its NP-hardness \cite{tillmann2014computational}.

@@

We begin by showing that $\dim(\bm{\mathcal{B}}_{\pi(S)}) = \dim(\bm{\mathcal{A}}_S)$ for all $S \in \mathcal{H}$. Since the right-hand side of \eqref{GapUpperBound} is less than one (by \eqref{delrho} and $r \leq 1$), it follows from \eqref{dimLem} that $|\pi(S)| \geq \dim(\bm{\mathcal{B}}_{\pi(S)}) \geq \dim(\bm{\mathcal{A}}_S) = |S|$, with the equality by injectivity of $\mathbf{A}$. Since $|S| = |\pi(S)|$, the required fact follows. Note that the columns of $\mathbf{B}_{\pi(S)}$ are therefore linearly independent, for all $S \in \mathcal{H}$.

We next show that \eqref{cond} holds.  Fix $\mathcal{G} \subseteq \mathcal{H}$. Since $\bm{\mathcal{B}}_{\cap \pi(\mathcal{G})} \subseteq \cap_{S \in \mathcal{G}} \bm{\mathcal{B}}_{\pi(S)}$, if $\cap_{S \in \mathcal{G}} \bm{\mathcal{B}}_{\pi(S)} = \{\textbf{0}\}$, then we must have $|\cap_{S \in \mathcal{G}} \pi(S)| = 0$ (as the columns of $\mathbf{B}_{\pi(S)}$ are linearly independent) and \eqref{cond} is trivially true. Suppose then that the intersection is not the zero vector. Keeping in mind that $d(U,V) \leq d(U',V)$ whenever $U \subseteq U'$ and applying (in order) Lem.~\ref{spanIntersectionLemma}, Lem.~\ref{DistanceToIntersectionLemma}, \eqref{eqdim}, \eqref{GapUpperBound} and finally \eqref{Cdef2} gives:
\begin{align}\label{randoml}
d(\bm{\mathcal{B}}_{\cap \pi(\mathcal{G})}&, \bm{\mathcal{A}}_{\cap \mathcal{G}}  ) 
\leq d\left( \cap_{S \in \mathcal{G}} \bm{\mathcal{B}}_{\pi(S)}, \cap_{S \in \mathcal{G}} \bm{\mathcal{A}}_S \right) \nonumber \\
&\leq \sum_{T \in \mathcal{G}} \frac{ d\left( \cap_{S \in \mathcal{G}} \bm{\mathcal{B}}_{\pi(S)},\bm{\mathcal{A}_{T}} \right) }{ r(  \{\bm{\mathcal{A}}_S\}_{S \in \mathcal{G}}) }
\leq \sum_{T \in \mathcal{G}} \frac{ d\left( \bm{\mathcal{B}}_{\pi(T)},\bm{\mathbf{A}}_{T} \right) }{ r( \{\bm{\mathcal{A}}_S\}_{S \in \mathcal{G}}) }\nonumber \\
&\leq \frac{|\mathcal{G}| \varepsilon}{r( \{\bm{\mathcal{A}}_S\}_{S \in \mathcal{G}})} 
\leq \frac{C_2 \varepsilon}{\max_i\|\mathbf{A}_i\|_2}. 
\end{align}
